\begin{doublespace}

\chapter*{Preface}

%\vspace{-50mm}

%{\Large Preface} \\

This book provides an introduction to statistics and its applications in the life sciences, and biomedical research.  It is based on the freely available \textsl{OpenIntro Statistics, Third Edition}, and, like \textsl{OpenIntro} it may be downloaded as a free PDF at \textbf{Need location}.  The text adds substantial new material, eliminates sections from \textsl{OpenIntro}, and re-uses some material directly.  Readers need not have read \textsl{OpenIntro}, since this book is intended to be used independently.  We have retained some of the exercises from \textsl{OpenIntro} that may not come directly from medicine or the life sciences but illustrate important ideas or methods that are commonly used in fields such as biology.

\textsl{Introduction to Statistics for the Life and Biomedical Sciences} is intended for graduate and undergraduate students interested in careers in biology or medicine, and may also be profitably read by students of public health.  It covers many of the traditional introductory topics in statistics used in those fields, but also adds some newer methods being used in molecular biology.  Statistics has become an integral part of research in medicine and biology, and the tools for displaying, summarizing and drawing inferences from data are essential both for understanding the outcomes of studies and for incorporating measures of uncertainty into that understanding.  An introductory text in statistics for students considering careers in medicine, public health or the life sciences should be more than simply thee usual introduction, supplemented with additional examples from biology or medical science. Along with the value of careful, robust analyses of experimental and observational data, it should convey some of the excitement of discovery that emerges from the interplay of science with data collection and analysis.  We hope we have conveyed some of that excitement here. 

\textit{DH: We have not conveyed that excitement so far.}

We have tried to balance the sometimes competing demands of mastering the important technical aspects of methods of analysis with gaining an understanding of important concepts.  The examples and exercises include opportunities for students to build skills in conducting data analyses and to state conclusions with clear, direct language that is specific to the context of a problem.  We also believe that computing is an essential part of statistics, just as mathematics was when computing was more difficult or expensive.  The text includes many examples where software is used to aid in the understanding of the features of a data as well as exercises where computing is used to help illustrate the notions of randomness and variability. Because they are freely available, we use the \textsf{R} statistical language with the \textsl{R Studio} interface.  Information on downloading \textsf{R} and \textsl{R Studio} is may be found in the Labs at \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}}.  Nearly all examples and exercises can be adapted to either SAS, Stata or other software, but we have not done that.

\textit{DH: This is where we should mention the parallel book of examples in R, reproducing the figures and tables in the book (without adornment) and providing a bit of instruction in R.}



\begin{comment}

This book may be downloaded as a free PDF at \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}}. %This download is free and need not be supplemented by the purchase of a paperback copy.
\vspace{3mm}

\noindent We hope readers will take away three ideas from this book in addition to forming a foundation of statistical thinking and methods.\vspace{-1mm}
\begin{enumerate}
\setlength{\itemsep}{0mm}
\item[(1)] Statistics is an applied field with a wide range of practical applications.
\item[(2)] You don't have to be a math guru to learn from real, interesting data.
\item[(3)] Data is messy, and statistical tools are imperfect. But, when you understand the strengths and weaknesses of these tools, you can use them to learn about the real~world.
\end{enumerate}
%This textbook is supplemented by many free, online resources to help students apply the methods they learn in this textbook and beyond.

\end{comment}

\subsection*{Textbook overview}

The chapters of this book are as follows:

\begin{description}
\setlength{\itemsep}{0mm}

%\begin{itemize}
\item[1. Introduction to data.] Data structures, variables, summaries, graphics, and basic data collection techniques.
\item[2. Probability.] The basic principles of probability. %An understanding of this chapter is not required for the main content in Chapters~\ref{modeling}-\ref{multipleAndLogisticRegression}.
\item[3. Distributions of random variables.] Introduction to the normal model and other key distributions.
\item[4. Foundations for inference.] General ideas for statistical inference in the context of estimating the population mean.
\item[5. Inference for numerical data.] Inference for one or two sample means using the normal model and $t$ distribution, and also comparisons of many means using ANOVA.
\item[6. Inference for categorical data.] Inference for proportions using the normal and chi-square distributions, as well as simulation and randomization techniques.
%\item[5. Large sample inference.] Inferential methods for one or two sample means and proportions using the normal model, and also contingency tables via chi-square.
%\item[6. Small sample inference.] Inference for means using the $t$ distribution, as well as simulation and randomization techniques for proportions.
\item[7. Introduction to linear regression.] An introduction to regression with two variables. Most of this chapter could be covered after Chapter~\ref{introductionToData}.
\item[8. Multiple and logistic regression.] An introduction to multiple regression and logistic regression for an accelerated course.

%\end{itemize}
\end{description}

\textbf{The remainder of this section requires revision}

%This textbook was written to allow flexibility in choosing and ordering course topics. The material is divided into two pieces: main text and special topics. The goal of the main text is to allow people to move towards statistical inference and modeling sooner rather than later. Special topics, labeled in the table of contents and in section titles, may be added to a course as they arise naturally in the curriculum.

\emph{OpenIntro Statistics} was written to allow flexibility in choosing and ordering course topics. The material is divided into two pieces: main text and special topics. The main text has been structured to bring statistical inference and modeling closer to the front of a course. Special topics, labeled in the table of contents and in section titles, may be added to a course as they arise naturally in the curriculum.

%\emph{OpenIntro Statistics} would also serve as a helpful supplement in a course preparing students for the Advanced Placement Statistics exam, either though the textbook or use of the online resources outlined below. %We hope for this textbook to be deemed ``AP ready'' for future editions, which will require an application for approval from College Board. %The AP material would include the main and optional materials of Chapters~1-7 with the possible omission of Sections~1.8, 3.5, 4.7, 6.3-6.4, and~7.4.

\subsection*{Examples, exercises, and appendices}

Examples and within-chapter exercises throughout the textbook may be identified by their distinctive bullets:

\begin{example}{Large filled bullets signal the start of an example.}
Full solutions to examples are provided and often include an accompanying table or figure.
 \end{example}

\begin{exercise}
Large empty bullets signal to readers that an exercise has been inserted into the text for additional practice and guidance. Students may find it useful to fill in the bullet after understanding or successfully completing the exercise. Solutions are provided for all within-chapter exercises in footnotes.\footnote{Full solutions are located down here in the footnote!}
\end{exercise}

There are exercises at the end of each chapter that are useful for practice or homework assignments. Many of these questions have multiple parts, and odd-numbered questions include solutions in Appendix~\ref{eoceSolutions}. %These end-of-chapter exercises are also available online in a public question bank at \textbf{openintro.org}, and the available selection is constantly growing based on teacher contributions. Numbered citations in end-of-chapter exercises may be found in Appendix~B.

Probability tables for the normal, $t$, and chi-square distributions are in Appendix~\ref{distributionTables}, and PDF copies of these tables are also available from \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}} for anyone to download, print, share, or modify.

\subsection*{OpenIntro, online resources, and getting involved}

OpenIntro is an organization focused on developing free and affordable education materials. \emph{OpenIntro Statistics}, our first project, is intended for introductory statistics courses at the high school through university levels.

%We encourage anyone learning or teaching statistics to visit \textbf{openintro.org} and get involved by using the many online resources, which are all free and rapidly expanding, or by creating new material. Students can test their knowledge with practice quizzes for each chapter, or try an application of concepts learned in each chapter using real data and the top-rated and free statistical software \emph{R}. Teachers can download the source for course materials, labs, slides, data sets, R figures, or create their own custom quizzes and problem sets for students to take on the website. Everyone is also welcome to download this textbook as a PDF or the book's source files to create a custom version of this textbook or to simply share a copy with a friend or on a website. All of these products are free, and we want to be clear that anyone is welcome to use these online tools and resources with or without this textbook as a companion.

We encourage anyone learning or teaching statistics to visit \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}} and get involved. We also provide many free online resources, including free course software. 
%, or by creating new material. Students can test their knowledge with practice quizzes for each chapter, or try an application of concepts learned using real data. 
Data sets for this textbook are available on the website and through a companion R package.\footnote{Diez DM, Barr CD, \c{C}etinkaya-Rundel M. 2012. \texttt{openintro}: OpenIntro data sets and supplement functions. \urlwofont{http://cran.r-project.org/web/packages/openintro}.} All of these resources are free, and we want to be clear that anyone is welcome to use these online tools and resources with or without this textbook as a companion.

%Teachers can download the source files for this book, labs, data sets, or create their own custom quizzes and problem sets for students to take at \href{http://www.openintro.org}{\textbf{openintro.org}}. 
%Anyone can download a PDF or the source files of this textbook for modifying and sharing at \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}}.

We value your feedback. If there is a particular component of the project you especially like or think needs improvement, we want to hear from you. You may find our contact information on the title page of this book or on the \href{http://www.openintro.org/about.php}{About} section of \href{http://www.openintro.org}{\color{black}\textbf{openintro.org}}.

\subsection*{Acknowledgements}

This project would not be possible without the dedication and volunteer hours of all those involved. No one has received any monetary compensation from this project, and we hope you will join us in extending a \emph{thank you} to all those volunteers below.

The authors would like to thank Andrew Bray, Meenal Patel, Yongtao Guan, Filipp Brunshteyn, Rob Gould, and Chris Pope for their involvement and contributions. %A special thank you to Andrew Bray, who has developed the R labs and diligently worked to translate end-of-chapter exercises to the project website. We also deeply appreciate the contribution of Meenal Patel, who has helped raise the professional profile of OpenIntro by designing a business system and website for the project. 
We are also very grateful to Dalene Stangl, Dave Harrington, Jan de Leeuw, Kevin Rader, and Philippe Rigollet for providing us with valuable feedback.

\textit{DH: We need to revise acknowledgements.}

% Add note about exercise/example circles

% Add note about margin notes for notation

\begin{comment}

@article{smith1998fifty,
  title={Fifty years of randomised controlled trials},
  author={Smith, R},
  journal={British Medical Journal},
  volume={317},
  pages={1166},
  year={1998}
}

\end{comment}

\end{doublespace}